---
title: "Calling Genotypes with and without a reference"
author: "William Hemstrom"
date: "2025-12-8"
output:
  html_document:
      toc: true
  pdf_document:
      toc: true
---

```{r setup}
library(data.table); library(foreach); library(ggplot2);library(snpR); library(sf)

add_path <- function(npath = "/home/hemstrow/bin:/home/hemstrow/miniforge3/bin:/usr/bin"){
  current_path <- Sys.getenv("PATH")
  Sys.setenv(PATH = paste(current_path, npath, sep = ":"))
}
```

# *de novo* genotyping (without a reference)

## Parameter control

The following R chunk controls the filtering parameters this script will use. To run this entire analysis with a different set of filters, they can simply be changed here! Everything that follows can then be re-run as written with no further changes necessary.

```{r set_filters}
# run parameters
par <- 4 # for all of these jobs, how many processing threads should we use?

# equalizing parameters
min_reads <- 2e4

# stacks parameters -- see below
# * M: number of mismatches allowed between stacks within individuals
# * m: minimum number of raw reads required to form a stack
# * n: number of mismatches allowed between stacks between individuals
m <- 4
M <- 2
n <- m
r <- .5

# ANGSD filters
minQ <- 20
mapQ <- 20
min_ind <- .5
maf <- 0.05

```

## Pre-processing

The first step is to filter our reads to remove low-quality bases, etc. We can do this for `stacks` specifically with `process_radtags`.

We're also going to run things in parallel using the `foreach` R package to run our processes on each sample at once.

```{r}
# make a list of our input files
SRA_info <- data.frame(
  raw = list.files("../fastqs/", ".+RA.+fastq$"))
SRA_info$R1 <- paste0("../fastqs/", SRA_info$raw)
SRA_info$R2 <- gsub("_RA_", "_RB_", SRA_info$R1)
SRA_info$sample <- gsub("_RA_", "_", SRA_info$raw)
SRA_info$sample <- gsub("SOMM371_", "", SRA_info$sample)
SRA_info$sample <- gsub("\\.fastq", "", SRA_info$sample)


trimmed_dir <- "./trimmed"
dir.create(trimmed_dir)

# set-up parallel computing
cl <- parallel::makePSOCKcluster(par)
doParallel::registerDoParallel(cl)

# run process_radtags for each sample!
foreach(q = 1:nrow(SRA_info)) %dopar% {
  add_path()
  
  # put together call
  cmd <- paste0("process_radtags ", 
                " -1 ", SRA_info[q, 2], # define read 1
                " -2 ", SRA_info[q, 3], # define read 2
                " -i fastq",
                " -y fastq",
                " -e SbfI",  # restriction enzyme
                " -o ", trimmed_dir, # output dest
                " --bestrad", # most of the time this isn't needed, quirk of this data
                " --disable-rad-check", # most of the time this isn't needed, quirk of this data
                " --truncate 110" # all reads need to be the same length for STACKS specifically
                )
  
  # run the call
  system(cmd)
}


parallel::stopCluster(cl)

SRA_info$R1_trimmed <- paste0("./trimmed/", SRA_info$sample, "_1.1.fq")
SRA_info$R2_trimmed <- paste0("./trimmed/", SRA_info$sample, "_2.2.fq")

# move files so the names are as expected
on <- gsub("\\.\\./fastqs/", "./trimmed/", SRA_info$R1)
on <- gsub("\\.fastq", "", on)
file.rename(paste0(on, ".1.fq"), SRA_info$R1_trimmed)
on <- gsub("_RA", "_RB", on)
file.rename(paste0(on, ".2.fq"), SRA_info$R2_trimmed)
```

## Equalizing read counts
Many analyses can be biased if downstream sequening depths are different. For reference guided assembly, we'd align to the reference at this point and equalize depths, but for *de novo* genotyping we'll instead equalize read counts here.

```{r}
# figure out read counts
system("pwd")
cmd <- paste0("wc -l ", SRA_info$R1_trimmed)
lens <- unlist(lapply(cmd, system, intern = TRUE)) # figure out the number of lines for each file
SRA_info$nreads <- as.numeric(gsub(" .+", "", lens))/4 # number of reads in each sample

ggplot(SRA_info, aes(x = log10(nreads))) +
  geom_histogram() +
  theme_bw() +
  geom_vline(xintercept = log10(min_reads), color = "red")

```

Pick `min_reads` and set above from this histogram.


```{r}
SRA_info$R1_sub <- paste0(SRA_info$sample, "_1_sub.fastq")
SRA_info$R2_sub <- paste0(SRA_info$sample, "_2_sub.fastq")

# subset
cl <- parallel::makePSOCKcluster(par)
doParallel::registerDoParallel(cl)
foreach(q = 1:nrow(SRA_info)) %dopar% {
  add_path()
  
  if(SRA_info$nreads[q] >= min_reads){
    # put together call -- R1
    cmd <- paste0("head -n ", min_reads*4, 
                  " ", SRA_info$R1_trimmed[q], # define read 1
                  " > ", SRA_info$R1_sub[q])
    
    # run the call
    system(cmd)
    
    
    # put together call -- R2
    cmd <- paste0("head -n ", min_reads*4, 
                  " ", SRA_info$R2_trimmed[q], # define read 1
                  " > ", SRA_info$R2_sub[q])
    
    # run the call
    system(cmd)
  }
}

parallel::stopCluster(cl)

SRA_info <- SRA_info[-which(SRA_info$nreads < min_reads),]
```

## Run STACKS

We're now ready to run stacks itself! This is a multiple step process. We'll run these tools in sequence:

1. ustacks: identify loci
2. cstacks: make a catalog
3. sstacks: match samples back to catalog
4. tsv2bam: transpose to bams (our actual per-sample 'alignments')
5. gstacks: call genotypes

We'll use the helper script 'denovo_map.pl' included with `stacks` to help do this for us. It needs two things:

1. a file noting the samples and their populations of origin (the 'popmap')
2. all of our samples in one directory, all of which should be zipped, with a specific naming convention.

First, we'll make the popmap file.

```{r}
meta1 <- fread("../meta/biosample.tsv")
meta2 <- fread("../meta/SRA_meadata.tsv")
meta <- merge(meta1, meta2)
meta <- merge(SRA_info, meta, by.y = "filename", by.x = "raw", all.x = TRUE, sort = FALSE)
meta$pop <- gsub("_S.+$", "", meta$sample_name)
meta$pop <- gsub("-", "_", meta$pop)
popmap <- data.frame(sample = paste0(SRA_info$sample),
                     pop = meta$pop)
write.table(popmap, "popmap.txt", col.names = FALSE, row.names = FALSE, quote = FALSE, sep = "\t")
```

Next, we'll rename and move the files. The files must all be named something like:
"sample.1.fq.gz" and "sample.2.fq.gz" for the read 1 and read 2 files.

```{r}
sampdir <- "./samples"
dir.create(sampdir)

# move files
SRA_info$R1_stacks <- paste0(sampdir, "/", SRA_info$sample, ".1.fq")
SRA_info$R2_stacks <- paste0(sampdir, "/", SRA_info$sample, ".2.fq")
file.rename(SRA_info$R1_sub, SRA_info$R1_stacks)
file.rename(SRA_info$R2_sub, SRA_info$R2_stacks)


# zip files
cl <- parallel::makePSOCKcluster(par)
doParallel::registerDoParallel(cl)
foreach(q = 1:nrow(SRA_info)) %dopar% {
  add_path()
  
  # zip
  cmd <- paste0("gzip ", SRA_info$R1_stacks[q])
  system(cmd)
  cmd <- paste0("gzip ", SRA_info$R2_stacks[q])
  system(cmd)
  
}
parallel::stopCluster(cl)
```

We're now ready to actually run stacks! This step uses three key parameters:

* M: number of mismatches allowed between stacks within individuals
* m: minimum number of raw reads required to form a stack
* n: number of mismatches allowed between stacks between individuals

```{r}
# create output dir
stdir <- "./stacks"
dir.create(stdir)

add_path()

cmd <- paste0("denovo_map.pl",
              " --samples ", sampdir,
              " --popmap ", "popmap.txt",
              " -o ", stdir,
              " -n ", n,
              " -M ", M,
              " -r ", r,
              " -T ", par,
              ' -X "ustacks: -m ', m, '"',
              ' -X "populations: --vcf"',
              ' -X "gstacks: --write-alignments"') # tells stacks to output a vcf file for further analysis

system(cmd)


on <- paste0("./stacks/", SRA_info$sample, ".alns.bam")
SRA_info$aln <- paste0("./alns/", SRA_info$sample, "alns.bam")
file.rename(on, aln)
```

# Do the low-coverage work including PCAngsd, etc.
Since this is *very* low coverage data, what we will actually do next is use the `catalog.fa.gz` reference and the `.aln.bam` alignments with ANGSD to get likelihoods from this. We don't expect many loci to be actually genotypable.

## parameters
```{r}
# filtering parameters handled above.
maxk <- 8
mapQ <- 20
minQ <- 20
snp_pval <- 1e-10
minInd <- .25
```

## re-align with bwa and filter via script
Since the alignments from STACKS don't include critical quality information that'll cause things to error down the line, we'll re-align using bwa.

```{r}
# index
system("bwa index stacks/catalog.fa.gz")
system("samtools faidx stacks/catalog.fa.gz")

cl <- parallel::makePSOCKcluster(par)
doParallel::registerDoParallel(cl)
foreach(q = 1:nrow(SRA_info)) %dopar% {
  add_path()
  cmd <- paste0("./run_align.sh ",
                SRA_info$R1_stacks[q], " ",
                SRA_info$R2_stacks[q], " ", 
               "stacks/catalog.fa.gz")
  system(cmd)
  
}
parallel::stopCluster(cl)

dir.create("alns")
SRA_info$aln <- paste0("alns/", SRA_info$sample, ".bam")
file.rename(paste0(gsub("\\.fq", ".sort.flt.bam", SRA_info$R1_stacks)),
            SRA_info$aln)
file.rename(paste0(gsub("\\.fq", ".sort.flt.bam.bai", SRA_info$R1_stacks)),
            paste0(SRA_info$aln, ".bai"))

```


## Get likelihoods
We use the "angsd.sh" script for this, which breaks things down into chunks to analyze. We start by defining region files using the index de-novo reference and making a bamlist.
```{r}
fai <- fread("stacks/catalog.fa.gz.fai")
fai[,r := paste0(V1, ":")]
rf <- split(fai$r, sort(rep(1:20, length.out = nrow(fai))))

dir.create("regionfiles")
for(i in 1:length(rf)){
  writeLines(rf[[i]], "rf_", i, ".txt")
}

bamlist <- writeLines(SRA_info$aln, "bamlist.txt")
```

We then run the script. This script runs for both maf 0.05 and maf 0.01.

```{r}
# bamlist
writeLines(SRA_info$aln, "bamlist.txt", sep = "\n")

cmd <- paste0("./angsd.sh ",
              mapQ, " ",
              minQ, " ",
              snp_pval, " ",
              floor(nrow(SRA_info)*minInd)
)

add_path("/home/hemstrow/miniforge3/envs/angsd/bin")
system(cmd)

like_res <- data.table(maf = c("01", "05"))
like_res[,beagle := paste0("darlingtonia_maf", maf, "_combined.beagle.gz")]
```

## run PCAngsd
Run PCAngsd on the results. Also calculates $F_{IS}$ (equivalent) per individual.

```{r}
dir.create("pcangsd")
for(i in 1:nrow(like_res)){
  cmd <- paste0("Rscript pcangsd_admix.R ",
                like_res$beagle[i], " ",
                "pcangsd/darlingtonia_maf", like_res$maf[i], "_pcangsd", " ",
                maxk, " ",
                par)
  
  add_path("/home/hemstrow/miniforge3/envs/pcangsd/bin")
  system(cmd)
}
```

Plot the results
```{r}
# prep
PCA_plots <- vector("list", nrow(like_res))
names(PCA_plots) <- like_res$maf
admix_plots <- PCA_plots
inbreed_plots <- admix_plots

popmap <- fread("popmap.txt", header = FALSE)
bamlist <- fread("bamlist.txt", header = FALSE)
colnames(bamlist) <- "bam"
colnames(popmap) <- c("sample", "pop")
bamlist[,sample := stringi::stri_extract(bam, regex = "[A-Z]+_[A-Z]+")]
bamlist <- merge(bamlist, popmap, by = "sample", sort = FALSE)

# plot
for(i in 1:nrow(like_res)){
  C <- as.matrix(read.table(paste0("pcangsd/darlingtonia_maf", like_res$maf[i], "_pcangsd.cov"))) # Reads estimated covariance matrix


  # Plot PCA plot
  e <- eigen(C)
  pca <- as.data.frame(e$vectors)
  colnames(pca) <- paste0("PC", 1:ncol(pca))
  pca$pop <- bamlist$pop
  pvarexp <- e$values/sum(e$values)
  
  PCA_plots[[i]] <- ggplot(pca, aes(x = PC1, y = PC2, color = pop)) +
    geom_point() +
    theme_bw() +
    khroma::scale_color_vibrant() +
    xlab(paste0("PC1 (", round(pvarexp[1]*100, 2), "%)")) +
    ylab(paste0("PC2 (", round(pvarexp[2]*100, 2), "%)")) +
    guides(color = guide_legend(title = "Population"))
  
  setwd("pcangsd")
  admix_plots[[i]] <- plot_structure(paste0("darlingtonia_maf", 
                                            like_res$maf[i], "_pcangsdK.+Q$"), 
                                     facet = bamlist$pop, 
                                     alt.palette = khroma::color("highcontrast")(2), k = 2)
  setwd("..")
  
  inbreed <- fread(paste0("pcangsd/darlingtonia_maf", like_res$maf[i], "_pcangsd.inbreed.samples"))
  colnames(inbreed) <- "FIS"
  inbreed$pop <- bamlist$pop
  inbreed_plots[[i]] <- ggplot(inbreed, aes(x = pop, y = FIS)) + 
    geom_boxplot() +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    xlab("Population") +
    ylab(bquote(italic(F[IS])))
}

```

Nothing really changes across MAFs, so we'll just show MAF = 0.05
```{r}
sp <- cowplot::plot_grid(PCA_plots$`05`, admix_plots$`05`$plot + xlab("Population"),
                   align = "hv", axis = "tlbr", nrow = 2)

cowplot::save_plot("../results/loco_structure.pdf", sp, base_height = 10, base_width = 10)
```

# IBS approach
This seems to work better with this very sparse data!
```{r}
cmd <- paste0("angsd -bam bamlist.txt -minQ 20 -minMapQ 20 -minMaf 0.05 -SNP_pval 1e-10 -minInd 45 -GL 1 -doMajorMinor 1 -doMaf 1 -doIBS 1 -doCounts 1 -doCov 1 -makeMatrix 1 -doGLF 2")

add_path()
system(cmd)

cov <- as.matrix(read.table("angsdput.covMat"))
cov <- na.omit(cov)
ommitted <- attr(cov, "na.action")
cov <- cov[,-ommitted]
e <- eigen(cov)
pca <- e$vectors
pca <- as.data.frame(pca)
colnames(pca) <- paste0("PC", 1:ncol(pca))
pca$pop <- popmap$pop[-ommitted]
percents <- e$values/sum(e$values)
pca <- as.data.table(pca)
pca[,Location := ifelse(pop == "Mendocino_County", "Mendocino",
                        ifelse(pop == "Plumas_NF", "Plumas",
                               ifelse(pop == "Shasta_Trinity_NF", "Shasta-Trinity",
                                      "Six Rivers")))]

PCA <- ggplot(pca, aes(x = PC1, y = PC2, color = Location)) + geom_point(size = 3) +
  khroma::scale_color_okabeito() + theme_bw() +
  xlab(paste0("PC1 (", round(percents[1], 3)*100, "%)")) +
  ylab(paste0("PC2 (", round(percents[2], 3)*100, "%)"))
```

# NGSadmix
Run
```{r}
k <- 1:4
reps <- 10

dir.create("NGSadmix")
for(i in k){
  for(j in 1:reps){
    output <- paste0("NGSadmix/darlingtonia_K", i, "_r", j)
    cmd <- paste0("~/bin/NGSadmix -likes angsdput.beagle.gz -K ", i, " -minMaf 0.05 -o ", output)
    system(cmd)
  }
}
```


```{r}
background <- rnaturalearth::ne_states(country = "United States of America", returnclass = "sf")
background <- background[background$name %in% c("Oregon", "Nevada", "California"),]
popmap[,Location := ifelse(pop == "Mendocino_County", "Mendocino",
                        ifelse(pop == "Plumas_NF", "Plumas",
                               ifelse(pop == "Shasta_Trinity_NF", "Shasta-Trinity",
                                      "Six Rivers")))]
meta <- as.data.table(meta)
popmap <- merge(popmap, meta[,.(sample, lat_lon)], by = "sample", sort = FALSE)
popmap[,Long := gsub(".+N ", "", lat_lon)]
popmap[,Long := gsub(" .+", "", Long)]
popmap[,Long := paste0("-", Long)]
popmap[,Lat := gsub(" .+", "", lat_lon)]
source("~/GITHUB/GRSP/scripts/genoscape_func.R")

ucoords <- unique(popmap[,.(Location, Long, Lat)])
ucoords$SG <- 1:nrow(ucoords)
ucoords <- ucoords[order(ucoords$Location),]
ucoords$SGn <- 1:nrow(ucoords)

popmap <- merge(popmap, ucoords[,.(Location, Long, Lat, SGn)], by = c("Location", "Long", "Lat"), sort = FALSE)
popmap[,n := .N, by = SGn]

# neword <- c(1,2,4,7, # six rivers
#             6, 8 , 9, 11, 12,# shasta
#             3, 5, # plumas,
#             10 # mendo
# )
# SG_key <- data.table(SG = 1:12)
# SG_key <- SG_key[neword,]
# SG_key[,newSG := 1:nrow(SG_key)]
# popmap[,SG := SG_key$newSG[match(popmap$group, SG_key$SG)]]


                     
Location <- popmap$Location
SG <- as.character(popmap$SGn)
setwd("NGSadmix/")
add_path()
admix_plots_full <- plot_structure("\\.qopt", Location, reps = 10, k = 2:4, clumpp_path = "~/bin/CLUMPP", 
                              alt.palette = khroma::color("okabeito")(4)[c(4,3,2,1)], qsort = FALSE, 
                              facet.order = c("Mendocino", "Plumas", "Shasta-Trinity", "Six Rivers"))
admix_plots_fine <- plot_structure("\\.qopt", SG, reps = 10, k = 2:4, clumpp_path = "~/bin/CLUMPP", 
                              alt.palette = khroma::color("okabeito")(4)[c(4,3,2,1)], qsort = FALSE, 
                              facet.order = as.character(c(1:14)))
setwd("..")
```

Map
```{r}
extent <- sf::st_bbox(background)
extent["ymin"] <- 38
extent["ymax"] <- 42.5
extent["xmax"] <- -119
background <- sf::st_crop(background, extent)

# veg data, from (Ecological Subsections: Landcover)[https://data.fs.usda.gov/geodata/edw/datasets.php?dsetCategory=biota]
veg <- st_read("~/GITHUB/rice_darlingtonia_2025/shpfiles/S_USA.NationalLandcoverSubsections/")
veg <- st_transform(veg, st_crs(background))
sf::sf_use_s2(FALSE)
veg <- st_crop(veg, extent)
veg$`% Forest Cover` <- veg$EVERGREEN1 + veg$DECID_FO_1 + veg$MIXED_FO_1

groups_large <- group_points(popmap, dist = .5, crs = st_crs(background), coords = c("Long", "Lat"), dist_is_degrees = TRUE)

psf <- st_as_sf(groups_large$simplified[,.(Location, Long, Lat)], coords = c("Long", "Lat"), crs = st_crs(background))

map <- ggplot()


map <- plot_structure_map(admix_plots_full, k = 3, facet = "Location", pop_coordinates = psf,
                                    layers = list(geom_sf(data = veg, aes(fill = `% Forest Cover`), color = NA, lwd = 0, alpha = .5),
                                                  khroma::scale_fill_batlow(reverse = TRUE),
                                                  # guides(fill = guide_legend(title = "% Forest Cover")),
                                                  geom_sf(data = background, fill = NA, lwd = 1, color = "black"),
                                                  ggnewscale::new_scale_fill()),
                                    radius_scale = .1, label_args = list(nudge_x = .5, nudge_y = .5),
                                    alt.palette = khroma::color("okabeito")(4)[c(4,3,2)], crop = FALSE,
                                    scale_bar = list(text_cex = 1, height = unit(.03, "native")),
                                    compass = list(style = ggspatial::north_arrow_orienteering, location = "br"), 
                                    pop_names = TRUE, ask = FALSE) +
  scale_x_continuous(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.05, 0.05))
map <- map + theme(legend.text = element_text(size = 14),
            axis.text = element_text(size = 12),
            axis.title = element_text(size = 16),
            legend.title = element_text(size = 16))
```

Final plot
```{r}
Figure_1 <- gridExtra::grid.arrange(PCA  + ggtitle("A") + theme(plot.title = element_text(size = 20)), 
                                    map + ggtitle("B")+ theme(plot.title = element_text(size = 20)), 
                                    admix_plots_full$plot + ylab("Ancestry Proportion") +
                                      ggtitle("C") + theme(plot.title = element_text(size = 20)),
                                    layout_matrix = matrix(c(1, 3, 2, 3), ncol = 2))
ggsave("Figure1.pdf", Figure_1, "pdf", height = 11, width = 15)


ggsave("FigureS1.pdf", admix_plots_fine$plot + ylab("Ancestry Proportion"), "pdf", height = 11, width = 15)
```



